\plaintitle

### 2025-04-23 docs/style/article-body.tex
added some details from initial experimentation

Toeplitz extraction been optimized quite well, and previous research can be utilized for this. However, there are still considerations when implementing the firmware for the microcontroller in order to optimize the code. Our goal is to attempt several implementations in order to find the most optimal implementation with the least amount of CPU-cycles.

@@

Initially, we will utilize MAX11102AUB\footnote{\href{https://www.farnell.com/datasheets/1913106.pdf}{Technical specification for MAX11102AUB, accessed 2025-03-13}} with an effective sample rate of 2 million samples per second (MSPS). This ADC provides a 12 bit sample size, providing roughly 2.86 MB/s of sampled data per second, derived by the following calculation.

@@

Many microcontrollers furthermore come equipped with internal ADCs that can be utilized, and while these provide a lower sample size (often around 1 MSPS), the ease of development may be prudent to utilize for this proof-of-concept. While our initial ADC has a fairly low throughput, this can always be upgraded if it ends up becoming too limiting. The final output from the ADC, whether built into the microcontroller or an external one, will be a stream of raw bits, as the analog signal from the OQRNG-device is processed.

@@

To summarize the theoretical working of Toeplitz extraction (as explained by Chouhan et al. \cite{toeplitz-desc}), the sampled raw bit matrix (\(T\)) are multiplied with a pre-determined seed matrix (\(K\)). The size of the seed is directly dependent on the size of the sampled data, and can be fixed or continually re-sampled as needed. To ensure high levels of entropy, our intuition is that re-sampling the seed from the OQRNG-device continually is prudent. The sample and seed with then be processed with matrix multiplication to remove deterministic patterns, and produce a bitstring that results in our randomly generated number. An example of how this extraction works can be seen below.

@@

With the consideration that our work revolves around optimizing Toeplitz extraction in order to quickly process random bits into a random number, we will take an iterative approach. For our tests, we will use a pre-defined stream of raw bits which is loaded into memory on the microcontroller, and run several different implementations of Toeplitz extraction to produce numbers. As we always use a pre-defined bitstream, the result will at this stage be deterministic, giving us a clear indication whether the algorithm works as intended.

@@

This naive implementation will then be flashed to our microcontrollers, beginning with Teensy 4.1 as this is the more capable of the microcontrollers used for this experiment. Code to measure the execution speed in milliseconds will be implemented and tested before we load the naive implementation on said microcontroller. We expect that several implementations may be too resource intensive or have a memory complexity far greater than our cheaper, less capable microcontroller are able to handle, and as such these may not be able to be tested until a few iterations of optimization has occurred.

@@

@@

Each iteration results in an implementation written in C, all of which will be supplied as an appendix to this paper. Furthermore, to simplify the implementation process, we will use the Arduino framework to communicate with the hardware. Whereas this introduces some overhead, it drastically reduces the time spent per iteration, and is deemed to be a suitable starting point. \newpage

@@

>>>>

Toeplitz extraction been optimized quite well, and previous research can be utilized for this. However, there are still considerations when implementing the firmware for the microcontroller in order to optimize the code. Our goal is to attempt several implementations in order to find the most optimal implementation with the least amount of effective processing time spent on the algorithm.

@@

Many microcontrollers furthermore come equipped with internal ADCs that can be utilized, and while these provide a lower sample size (often around 1 MSPS), the ease of development may be prudent to utilize for this proof-of-concept. While our initial ADC has a fairly low throughput, this can always be upgraded if it ends up becoming too limiting.

Should these internal ADCs prove too limiting, we propose utilizing MAX11102AUB\footnote{\href{https://www.farnell.com/datasheets/1913106.pdf}{Technical specification for MAX11102AUB, accessed 2025-03-13}} with an effective sample rate of 2 million samples per second (MSPS). This ADC provides a 12 bit sample size, providing roughly 2.86 MB/s of sampled data per second, derived by the following calculation.

@@

The final output from the ADC, whether built into the microcontroller or an external one, will be a stream of raw bits, as the analog signal from the OQRNG-device is processed.

@@

To summarize the theoretical working of Toeplitz extraction (as explained by Chouhan et al. \cite{toeplitz-desc}), the sampled raw bit matrix (\(T\)) is multiplied with a pre-determined seed matrix (\(K\)). The size of the seed is directly dependent on the size of the sampled data, and can be fixed or continually re-sampled as needed. To ensure high levels of entropy, our intuition is that re-sampling the seed from the OQRNG-device continually is prudent. The sample and seed with then be processed with matrix multiplication to remove deterministic patterns, and produce a bitstring that results in our randomly generated number. An example of how this extraction works can be seen below.

@@

With the consideration that our work revolves around optimizing Toeplitz extraction in order to quickly process random bits into a random number, we will take an iterative approach. For our tests, we will use a pre-defined stream of raw bits which is sent to the microcontroller via USB, and run several different implementations of Toeplitz extraction to produce numbers. As we always use a pre-defined bitstream, the result will at this stage be deterministic, giving us a clear indication whether the algorithm works as intended.

@@

This naive implementation will then be flashed to our microcontrollers, beginning with Teensy 4.1 as this is the more capable of the microcontrollers used for this experiment. Code to measure the execution speed in microseconds will be implemented and tested before we load the naive implementation on said microcontroller. We expect that several implementations may be too resource intensive or have a memory complexity far greater than our cheaper, less capable microcontroller are able to handle, and as such these may not be able to be tested until a few iterations of optimization has occurred.

@@

In order to evaluate the efficacy of each iteration, we will test the code by processing our pre-determined bitstreams, saving the processed bits in a binary file and evaluating this file with the command line utility \texttt{ent}. This utility runs a series of tests on the data to determine its level of entropy the file has -- rather, how information dense the file is. A perfect result (\emph{e.g. entirely random}) has an entropy value of \(8.0\), whereas acceptable entropy values for our tests are greater than \(7.99\), denoting a high level of randomness. Furthermore, \texttt{ent} can verify the arithmetic mean of the file, summing all the bits in the file and dividing it by the length of the file. A good distribution of randomness should have an arithmetic mean of \(~127.5\). To ensure each iteration works as expected, we will compare the \texttt{ent} result from each iteration against the original baseline, ensuring that the resulting processed bitstring is identical between iterations. There are a few more measurements shown by \texttt{ent}, and more details regarding these can be found in the utilities manual page\footnote{\href{https://manpages.ubuntu.com/manpages/trusty/man1/ent.1.html}{Manual page for \texttt{ent}, accessed 2025-04-23.}}.

@@

Each iteration results in an implementation written in C++, all of which will be supplied as an appendix to this paper. Furthermore, to simplify the implementation process, we will use the Arduino framework to communicate with the hardware. Whereas this introduces some overhead, it drastically reduces the time spent per iteration, and is deemed to be a suitable starting point.

Initially, we created a script to simulate a stream of bits from a file, and pre-computed the expected results using a trusted version of Toeplitz extraction. With the expected results saved to a file on the host computer, each iteration could then be deployed to the MCU, and tested with the same data as the baseline. As the algorithm is fully deterministic with this uniform data, the exact same results in the same order should be produced by all iterations. The simulation script then simply asserts that the MCU produces the same random string of bits from the resulting iterations as with the baseline.

As our system does not rely on interrupts in any capacity, measuring the execution speed of the implemented Toeplitz extraction function can be done using the built in \texttt{micros()}-function provided by Arduino. This provides us with an accurate measurement of the execution time per function call. The average execution time per test set over each iteration will be presented in \textbf{\texttt{TABLE\ 1}}, along with the fastest and slowest executions per iteration respectively. Furthermore, the entropy value on the processed bitstream will be noted in the table to demonstrate that the important measurement, e.g.~the randomness, still holds true between iterations.

\newpage

@@

2025-04-22: Begun including details regarding initial experimentation, updating details regarding experiments that had to change (e.g.~no baseline on separate hardware).
